---
title: "GSS challange"
author: "Anjolie Barrios"
date: "4 March 2021"
output: html_document
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(knitr)
library(orsk)
GSScsv <- read_csv("GSScsv.csv", col_types = cols(VOTE16 = col_factor(levels = c("Voted", "Did not vote")), RACE = col_factor(levels = c("BLACK", "WHITE", "OTHER"))))
GSScsv <- GSScsv 
  EVOLVED <- as.factor(GSScsv$EVOLVED)
```

# Introduction

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?.  If you skimp in this section expect to be marked down in your grade. Paint a picture of the project a couple of paragraphs is expected!-->

Is belief in the Big Bang Theory (BIGBANG) independent of race (RACE)?
If not, then whose view differs from whom? 

$H_0$ Belief in the Big Bang Theory has no association with race.  
  
$H_A$ Belief in the Big Bang Theory is not independent of race; in general, one or more races differ from another in this aspect.  

# Methods  
RACE vs BIGBANG, RACE vs EVOLVED, and BIGBANG vs DEGREE are the most important comparisons to be made in the data for this write-up. RACE is the case's race, categorized as black, white, or "other; DEGREE is the case's highest educational degree obtained (from light high school to college graduate); BIGBANG is whether they believe in the Big Bang Theory or not; EVOLVED is whether they believe in the theory of evolution.  
<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method. Discuss what data manipulations may be needed for example subsetting or other binning procedures and what sort of graphical displays and inferential tools will be used!-->

# Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

## Descriptive Results

### Graphical Descriptive Results
```{r}
gss_chi <- GSScsv %>% group_by(RACE,BIGBANG) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))
basicC <- ggplot(gss_chi,aes(x=RACE,y=count,fill=BIGBANG))
basicC + geom_bar(stat="identity",position="dodge")

basicCC <- ggplot(gss_chi,aes(x=RACE,y=prcnt*100,fill=BIGBANG)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```
  
The white and "other" races have similar amounts of people who believe and disbelieve in the Big Bang Theory, but by percent the "other" races have a tendency to believe in it. Out of the blacks interviewed, however, more than 4 times as many who believed were disbelievers.  
```{r}
gss_raceDegree <- GSScsv %>% group_by(RACE,DEGREE) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))
basicC <- ggplot(gss_raceDegree,aes(x=RACE,y=count,fill=DEGREE))
basicC + geom_bar(stat="identity",position="dodge")

basicCC <- ggplot(gss_raceDegree,aes(x=RACE,y=prcnt*100,fill=DEGREE)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```
  
The other races have an equal number of bachelor, high school, and LT high school students. However, a disproportionate number of high-schoolers were interviewed. Whites, by percent, were more likely to be in the other stages of education than blacks. 
```{r}
gss_bigbangDegree <- GSScsv %>% group_by(DEGREE,BIGBANG) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))
basicC <- ggplot(gss_bigbangDegree,aes(x=DEGREE,y=count,fill=BIGBANG))
basicC + geom_bar(stat="identity",position="dodge")

basicCC <- ggplot(gss_bigbangDegree,aes(x=DEGREE,y=prcnt*100,fill=BIGBANG)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```
  
Graduates and bachelors were slightly more likely to believe in the Big Bang than those in lower education levels.  
```{r}
gss_raceEvolved <- GSScsv %>% group_by(RACE,EVOLVED) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))
basicC <- ggplot(gss_raceEvolved,aes(x=RACE,y=count,fill=EVOLVED))
basicC + geom_bar(stat="identity",position="dodge")

basicCC <- ggplot(gss_raceEvolved,aes(x=RACE,y=prcnt*100,fill=EVOLVED)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```  
  
Blacks and whites are about even, but the "other" races are more likely to think the theory of evolution is false.  
<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question. Bottom line is to explain the features of the graph that are relevant to conforming or departing to what is expected by your null hypothesis !-->

### Numerical Descriptive Results
```{r}
table1 <- xtabs(~RACE + BIGBANG, data=GSScsv)
rowPerc(table1)
colPerc(table1)
```  
80.77% of blacks thought the Big Bang was false, while only 19.23% believed it. Whites are over-represented in this survey, so it doesn't really matter that 21.88% of the people who didn't believe were black. 
```{r}
table2 <- xtabs(~RACE + DEGREE, data=GSScsv)
rowPerc(table2)
colPerc(table2)
```  
High- schoolers were over-represented, especially in the black category; 30.77% of the blacks finished some college (graduate or bachelor), but 38.06% of the whites finished some college. Thus, while bachelors and graduates were overrepresented by whites (like every other category), race doesn't seem to be associated with educational level in any significant way.  
```{r}
table3 <- xtabs(~DEGREE + BIGBANG, data=GSScsv)
rowPerc(table3)
colPerc(table3)
```  
Higher education levels are slightly correlated with believing in the Big Bang. 
```{r}
table4 <- xtabs(~RACE + EVOLVED, data=GSScsv)
rowPerc(table4)
colPerc(table4)
```
Blacks are exactly even, but 2/3 of the "other" races don't believe in the theory of evolution. White is mostly even.  
```{r}
table5 <- xtabs(~RACE + POSTLIFE, data=GSScsv)
rowPerc(table5)
colPerc(table5)
```
"Yes" in Postlife is more common than "no" for all races about evenly.  
<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

## Inferential Results
```{r}
#race vs EVOLVED
chisqtestGC(table4)
#race vs BIGBANG
chisqtestGC(table1)
#degree vs BIGBANG
chisqtestGC(table3)
```
  
The null hypothesis is that BIGBANG, EVOLVED, race, and DEGREE have no association with each other. BIGBANG, when compared to race and highest scholastic degree, has a higher Chi-Square Statistic (and thus differs more than expected from the null hypotheses) than RACE vs EVOLVED does. However, only RACE vs BIGBANG has a p-value (of 0.011) low enough confirm the alternate hypothesis.  
  
### Fischer Tests  
This defines the FischerXY function.    
```{r}
fisherXY = function(x, y, confL = 0, printAll=FALSE){
  # Inputs: 
  #   x: factor variable, 2 levels
  #   y: factor variable, n levels
  # confL: confidence level, default = 0.95
  #        print only those tables, results if p-value < (1-confL)
  # printAll: overide the 'print only if p-value < (1-confL), and
  #           print all tables and fisher test results
  #
  # Constraint:
  #   x and y must be same length
  #
  # Outputs:
  #   All 2x2 Fisher exact tests
  
  if(confL<0 || confL> 1){
    print('error: confL must be between 0 and 1')
    return(999)
  }
  
  if(!is.factor(x)) x=as.factor(x)
  if(!is.factor(y)) y=as.factor(y)
  
  fisherTable = as.matrix(table(x, y))
  cat('\ncontingency table:\n')
  print(fisherTable)
  cat('\n***All 2X2 tests****\n\n')

  # all 2x2 tests
  levelsY = levels(y)
  levelsX = levels(x)
  
  m = length(levelsX)
  n = length(levelsY)
  
  confLevel = 0.95
  if (confL !=0 ) confLevel=confL
  
  for (i in 1:(m-1))
       for(j in ((i+1):m))
          for (k in 1:(n-1))
             for (l in ((k+1):n)) {
                
                 fisherIndexX = c(i,j)
                 fisherIndexY = c(k,l)
                 
                 fTempTable = as.matrix(fisherTable[fisherIndexX,
                                        fisherIndexY])
                 
                 fisherTest=fisher.test(fTempTable, conf.int = TRUE, 
                              conf.level = confLevel)
                 
                 if((fisherTest$p.value < (1-confLevel) ) || printAll){
                   
                       print(c(levelsX[i], levelsX[j]))
                       print(c(levelsY[k], levelsY[l]))
                       print(fTempTable)
                       print(fisherTest)
                       cat('\n')
                 }
                 
        }
}

```

#### Race vs BIGBANG
```{r}
fisherXY(GSScsv$RACE,GSScsv$BIGBANG,confL=.95,printAll = T)
```
  
A p-value of 0.8029 means that, assuming WHITE and OTHER have no difference in BIGBANG (the null hypothesis), there's an 80.29% chance that the data will be exactly as it is. Comparisons between blacks and whites/others have p-values much lower than 0.5 and thus confirm the alternate hypothesis: that race is associated with belief in the Big Bang theory.  
White and Other have the smallest differences in beliefs; 1.2482 is the odds ratio there.  
When comparing blacks and whites (with White as the control group), the risk ratio is ~162%, meaning Blacks are 1.62 times as likely to disbelieve in the Big Bang Theory than whites.  
Comparing whites vs Other (White as control group), the RR is ~88.9%, meaning people who are neither black or white are only 0.88 times as likely to disbelieve than whites.  
Finally, comparing blacks vs Other, the RR is ~55.03%; those in the "Other" category are more likely to believe in the theory than blacks. 

#### Race vs EVOLVED
```{r}
fisherXY(GSScsv$RACE,GSScsv$EVOLVED,confL=.95,printAll = T)
```
  
With p-values of 1, blacks and whites don't significantly differ in their beliefs in evolution.  
Black vs Other and White vs Other have odds ratios that are close to each other and much less than 1 (at around 0.5), which makes sense given that the graphs show that those of Other races have more of a tendency to disbelieve in evolution.  

#### Degree vs BIGBANG
```{r}
fisherXY(GSScsv$DEGREE,GSScsv$BIGBANG,confL=.95,printAll = T)
```
  
The Bachelor vs Graduate comparison has odds ratios close to 1, meaning that their beliefs in the Big Bang are similar, leaning towards TRUE. Comparisons between LT High School, High School, and Junior College also have odds ratios close to 1, leaning towards FALSE. Bachelor & Graduate will have odds ratios of less than 0.6 when compared to any of the other three.  
<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->


# Discussion and Conclusion
Race and belief in the Big Bang are associated, confirming the alternate hypothesis; blacks are far more likely to disbelieve in it (whites and Others only have small differences).  
This is not caused, necessarily, by increased religiosity nor education levels; education levels aren't associated closely with race, and the races have very different reactions to evolution and the Big Bang Theory (both religiously charged topics). Thus, there isn't enough information to properly explain why the alternate hypothesis was true.  

<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->
